\bfseries Convolution-based functions on tensor fields and tensor images

### 2022-08-19 paper/main.tex

@@

@@

% typesetting adjustments
\makeatletter
\renewcommand\section{\@startsection {section}{1}{\z@}%
  {-3.25ex \@plus -1ex \@minus -.2ex}%
  {1.5ex \@plus .2ex}%
  {\raggedright\normalfont\large\bfseries}}
\makeatother
\setlength{\textwidth}{5.00in}
\setlength{\oddsidemargin}{3.25in}
\addtolength{\oddsidemargin}{-0.5\textwidth}
\setlength{\textheight}{9.40in}
\setlength{\topmargin}{-0.50in}
\setlength{\footskip}{0.25in}
\pagestyle{myheadings}
\markboth{foo}{Hogg, Blum-Smith, \& Villar /Geometric filters for geometric images}
\newcommand{\figurerule}{\rule[1ex]{\textwidth}{0.2pt}}
\linespread{1.08}
\frenchspacing\sloppy\sloppypar\raggedbottom

\title{\bfseries%
Equivariant convolution-based functions
on tensor fields}
\author{David W. Hogg, Ben Blum-Smith, and Soledad Villar}
\date{}

@@

\paragraph{Abstract:} We provide a construction for general equivariant convolution-based functions of images or lattices
containing scalars, vectors, and tensors. The formulation, inspired by Einstein summation notation, does not use point-wise activation functions, but it can universally approximate all analytic equivariant functions with respect to discrete rotations of the image. CHECK WITH BEN

@@

Contemporary physics teems with data sets that are images or lattices or grids of geometric objects.
Rectangular images or lattices of density (a scalar field), temperature (also a scalar field), velocity (a vector field), magnetic field (a pseudo-vector field), and polarization (a 2-tensor field) are regularly taken with detectors or made in simulations.
That is, physics has lots of data that consists of geometric objects on regular grids.
These objects (scalars, vectors, pseudo-vectors, and tensors, for example) are \emph{geometric} in the sense that they are defined by their transformation properties under rotation and reflection.
For example, scalars are defined to be quantities in $\mathbb R$ that are invariant with respect to all rotations and reflections; vectors are defined (in $d$-dimensional space) to be quantities in $\mathbb R^d$ that transform under rotation or reflection by multiplication by the orthogonal $d\times d$ matrix representing the rotation or reflection; and so on.

For one concrete example, cosmological n-body simulations comprise the most computationally expensive part of most contemporary cosmological projects, missions, and experiments (CITE THINGS).
For this reason, accurate, fast emulators might be of enormous economic and scientific value.
There are now quite a few successful projects implementing such emulators (CITE THINGS).
Importantly, the initial conditions for these simulations are tiny displacements (vectors) and velocities (vectors) relative to a cubical lattice on a 3-torus.
That is, the initial conditions can be described as a 3-dimensional image of vectors on a regular grid.
For another concrete example, astronomical imaging data taken of the interstellar medium can be represented as rectangular lattices of measurements of temperature (scalar), optical depth (scalar), and induced polarization of starlight (a 2-tensor).
Here the data can be described as a 2-dimensional image of scalars and tensors on a regular grid.

@@

\begin{figure}[tp]
\includegraphics[width=\textwidth]{notebooks/filter+0_2_3.png}\\
\includegraphics[width=\textwidth]{notebooks/filter+1_2_3.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-1_2_3.png}
\caption{All the filters for $d=2$, $m=1$. HOGG: notes!}
\end{figure}

\begin{figure}[tp]
\includegraphics[width=\textwidth]{notebooks/filter+0_2_5.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-0_2_5.png}\\
\includegraphics[width=\textwidth]{notebooks/filter+1_2_5.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-1_2_5.png}
\caption{All the filters for $d=2$, $m=2$. HOGG: notes!}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\textwidth]{notebooks/monomials_1.png}
    \caption{A scalar image $s$ and convolutions with four chosen filters.}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\textwidth]{notebooks/monomials_2.png}
    \caption{All unconvolved second-degree monomials of $s$, given four chosen filters. For the 2-tensor outputs (order $k=2$), we traced them back to scalars.}
\end{figure}

>>>>

% write the levi civita contraction
% write the procedure of which we make all linear functions from k tensor images to k' tensor images of the same size. 
% do we include a pooling move?

% how do we enumerate all tensor functions

% 1, 6, 22, 49, 87 d=3, n=1,3,7,9 (11N^2-24N+21)/8

@@

% stuff for framing figures
\usepackage[framemethod=tikz]{mdframed}
\usetikzlibrary{shadows}
\definecolor{captiongray}{HTML}{555555}
\mdfsetup{%
  innertopmargin=2ex,
  innerbottommargin=1.8ex,
  linecolor=captiongray,
  linewidth=0.5pt,
  roundcorner=5pt,
  shadow=true,
  shadowcolor=black!05,
  shadowsize=4pt
}

@@

% typesetting adjustments
\makeatletter
\renewcommand\section{\@startsection {section}{1}{\z@}%
  {-3.25ex \@plus -1ex \@minus -.2ex}%
  {1.5ex \@plus .2ex}%
  {\raggedright\normalfont\large\bfseries}}
\makeatother
\setlength{\textwidth}{5.00in}
\setlength{\oddsidemargin}{3.25in}
\addtolength{\oddsidemargin}{-0.5\textwidth}
\setlength{\textheight}{9.40in}
\setlength{\topmargin}{-0.50in}
\setlength{\footskip}{0.25in}
\pagestyle{myheadings}
\markboth{foo}{Hogg, Blum-Smith, \& Villar / Geometric filters for geometric images}
\newcommand{\figurerule}{\rule[1ex]{\textwidth}{0.2pt}}
\linespread{1.08}
\frenchspacing\sloppy\sloppypar\raggedbottom

\title{\bfseries%
Convolution-based functions
on tensor fields and tensor images}
\author{David W. Hogg, Ben Blum-Smith, and Soledad Villar}
\date{}

@@

\paragraph{Abstract:}
Convolutional neural networks are the main tools for machine learning with images at the present day.
These methods effectively assume that the input images are scalar images (an intensity or brightness in each pixel, possibly in three or more channels), and they enforce a kind of locality and translation-equivariance.
In natural-science domains, image-like data sets can have a vector (velocity, say), or tensor (polarization, say), or pseudovector (magnetic field, say), or any combination in each pixel.
We provide a construction for general convolution-based functions of images or lattices
containing scalars, vectors, and tensors, producing scalar, vector, and tensor outputs.
Our formulation, which combines convolutions with products, tensor contractions, and tensor index permutations, does not require point-wise activation functions (RELU, say) but can universally approximate any suitably local analytic scalar, vector, or tensor functions on scalar, vector, or tensor images.
Our method permits, with a very simple adjustment, restriction to function spaces that are exactly equivariant to translations, discrete rotations, and reflections.

@@

Contemporary natural science teems with data sets that are images or lattices or grids of geometric objects.
Rectangular images or lattices of density (a scalar field), temperature (also a scalar field), velocity (a vector field), magnetic field (a pseudo-vector field), and polarization (a 2-tensor field) are regularly taken with detectors or made in simulations.
That is, there is plenty of data consisting of geometric objects on regular grids.
These kinds of images (grids of scalars, vectors, pseudo-vectors, and tensors, for example) are \emph{geometric} in the sense that they contain (at every pixel) objects are defined by their transformation properties under rotation and reflection.
For example, scalars are defined to be quantities in $\mathbb R$ that are invariant with respect to all rotations and reflections; vectors are defined (in $d$-dimensional space) to be quantities in $\mathbb R^d$ that transform under rotation or reflection by multiplication by the orthogonal $d\times d$ matrix representing the rotation or reflection; and so on.

For a concrete example close to one of our hearts (DWH's), contemporary cosmological n-body simulations comprise the most computationally expensive part of most contemporary cosmological projects, missions, and experiments (CITE THINGS).
For this reason, accurate, fast emulators might be of enormous economic and scientific value.
There are now quite a few successful projects implementing such emulators (CITE THINGS).
Importantly, the initial conditions for these simulations are tiny displacements (vectors) and velocities (vectors) relative to a cubical lattice on a 3-torus.
That is, the initial conditions can be described as a 3-dimensional image of vectors on a regular grid.
For another concrete example, astronomical imaging data taken of the interstellar medium can be represented as rectangular lattices of measurements of temperature (scalar), optical depth (scalar), and induced polarization of starlight (a 2-tensor).
Here the data can be described as a 2-dimensional image of scalars and tensors on a regular grid.

@@

\begin{figure}[tp]
  \begin{mdframed}
  \color{captiongray}
  \begin{center}
\includegraphics[width=\textwidth]{notebooks/filter+0_2_3.png}\\
\includegraphics[width=\textwidth]{notebooks/filter+1_2_3.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-1_2_3.png}
\includegraphics[width=\textwidth]{notebooks/filter+2_2_3.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-2_2_3.png}
  \end{center}
\caption{All the filters for $d=2$, $m=1$. HOGG: notes!}
  \end{mdframed}
\end{figure}

\begin{figure}[tp]
  \begin{mdframed}
  \color{captiongray}
  \begin{center}
\includegraphics[width=\textwidth]{notebooks/filter+0_2_5.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-0_2_5.png}\\
\includegraphics[width=\textwidth]{notebooks/filter+1_2_5.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-1_2_5.png}
  \end{center}
\caption{All the filters for $d=2$, $m=2$. HOGG: notes!}
  \end{mdframed}
\end{figure}

@@

\begin{figure}
  \begin{mdframed}
  \color{captiongray}
  \begin{center}
    \includegraphics[width=\textwidth]{notebooks/monomials_1.png}
  \end{center}
    \caption{A scalar image $s$ and convolutions with four chosen filters.}
  \end{mdframed}
\end{figure}

@@

\begin{figure}
  \begin{mdframed}
  \color{captiongray}
  \begin{center}
    \includegraphics[width=\textwidth]{notebooks/monomials_2.png}
  \end{center}
    \caption{All unconvolved second-degree monomials of $s$, given four chosen filters. For the 2-tensor outputs (order $k=2$), we traced them back to scalars.}
  \end{mdframed}
\end{figure}
